---
title: "Reproducible Research: Peer Assignment 1"
author: "Russell Dinnage"
date: "13/06/2014"
output: html_document
---

# Reproducible Research: Peer Assessment 1

This report describes an analysis of some data from a personal monitoring device, that has recorded the number of steps that an anonymous individual has taken in 5-minute intervals over two months in 2012.

## Loading and preprocessing the data

The first thing to do is load the data into R.

```{r, results='asis'}
## load required packages
library(knitr)
dat <- read.csv("activity.csv")
## check that the data loaded correctly
kable(head(dat))
```

That looks ok, so we move on to the next question. 

## What is mean total number of steps taken per day?

I am going to use the package `dplyr` to do most of this data manipulation, so I will load that up first. Then I will generate a histogram for the number of steps taken per day in this data set.

```{r}
library(dplyr)
library(wesanderson) ## this is just for fun (it's a palette package :)
## sum up all steps by date
steps_by_day <- dat %>% group_by(date) %>% summarise(Steps = sum(steps, na.rm = TRUE)) 
## plot histogram
hist(steps_by_day$Steps, breaks = 12, col = wes.palette(5, "Darjeeling"), xlab = "Number of Steps per Day", main = "")
```

As one might expect, the distribution is fairly Gaussian in shape except for the spike at zero. This is probably because some of the days were all NAs and the sum function will return zero if it removes all of the values because they were all NA. The mean number of steps per day was `r mean(steps_by_day$Steps)` and the median was `r median(steps_by_day$Steps)`. Those values were calculated using:

```{r}
mean(steps_by_day$Steps)
median(steps_by_day$Steps)
```








## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
